The Magnetorotational Instability Prefers Three Dimensions

### 2019-07-23 paper/paper.tex
Discussion of results

We solve the linearized magnetohydrodynamic equations in a frame rotating with angular frequency $\Omega$. The Navier-Stokes equation is in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate.
It is common in the accretion disk literature to relate $S$ and $f$ via the $q = S/f$ parameter; for all runs here, $q = 0.75$.
$Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $L_x = \pi$. The walls at $x = \pm L_x/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

@@

Our most surprising result is that at small values of supercriticality, the fastest growing mode is not axisymmetric. 
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate is found at $(k_y, k_z) \simeq (0.263, 0.447)$.
By expanding equations~(\ref{eq:ns}-~\ref{eq:divu}) in terms of $S = \SSC + \epsilon R$, we compute the leading order correction to the growth rate from three-dimensional effects as
\begin{widetext}
\begin{equation}
  \label{eq:asymp}
\omega^{2} \ = \ \frac{q\,B^{2}}{(q+1)} \left[ k_z^2 \left(\frac{d^2}{\pi^{2}} k_z^2-R\right) \ - \ \frac{\left(6+\pi ^2\right) q+\pi
   ^2-6}{12 } \, k_{y}^{2}  \right] \ + \ ...
\end{equation}
\end{widetext}
Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 <= k_y < 0.2$ to this asymptotic approximation, showing that the numerical calculations accurately reproduce the analytic ones in the region where the latter is valid.
All unstable modes are purely growing (figure~\ref{fig:growth_rate}c); the stable modes are rotationally modified Alfven waves found in left- and right- going pairs, consistent with the analytic predictions for two-dimensional stability calculations.
%
\onecolumngrid

>>>>

We solve the linearized magnetohydrodynamic equations in a frame rotating with angular frequency $\Omega$. The Navier-Stokes equation is in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate.
It is common in the accretion disk literature to relate $S$ and $f$ via the $q = S/f$ parameter; for all runs here, $q = 0.75$.
$Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $d = \pi$. The walls at $x = \pm d/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

@@

Our main result is that at small values of supercriticality, the fastest growing mode is not axisymmetric. 
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate is found at $(k_y, k_z) \simeq (0.263, 0.447)$.
In fact, at $\SSC = 0.640$, there are no unstable two-dimensional modes, but three-dimensional instability is present.

While surprising, this result can be predicted analytically.
By expanding equations~(\ref{eq:ns}-\ref{eq:divu}) in terms of $S = S_c + \epsilon R$, we compute the leading order correction to the growth rate from three-dimensional effects as
\begin{widetext}
\begin{equation}
  \label{eq:asymp}
\sigma^{2} \ = -\frac{q\,B^{2}}{(q+1)} \left[ k_z^2 \left(\frac{d^2}{\pi^{2}} k_z^2-R\right) \ - \ \frac{\left(6+\pi ^2\right) q+\pi
   ^2-6}{12 } \, k_{y}^{2}  \right] \ + \ ...
\end{equation}
\end{widetext}
The first term in equation~(\ref{eq:asymp}) is the result from a two-dimensional calculation.
The second term is negative definite and proportional to $k_y^2$: for sufficiently large $k_y$, it leads to instability even when the two-dimensional modes are stable.

Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 \le k_y \le 0.2$ to this asymptotic approximation, showing that the numerical calculations accurately reproduce the analytic ones in the region where the latter is valid.
We provide the full spectrum for the most unstable parameter values for $\SSC = 1.02$ in figure~\ref{fig:growth_rate}c.
The single unstable mode is purely growing (orange dot), and its complex conjugate is also present (blue dot at $\omega = 0$), consistent with equation~(\ref{eq:asymp}).
The other stable modes (blue dots) are rotationally modified Alfven waves found in left- and right- going pairs, consistent with the analytic predictions for two-dimensional stability calculations.
%
\onecolumngrid
